\documentclass[a4paper,12pt]{letter}

% Some of the article customisations are relevant for this class

\name{Grant R. Wythoff} % To be used for the return address on the envelope
\signature{Grant R. Wythoff} % Goes after the closing (ie at the end of the letter, with space for a signature)
\address{Grant Wythoff \\ Postdoctoral Fellow \\ Society of Fellows in the Humanities \\ Lecturer, Department of English \\ Columbia University}
% Alternatively, these may be set on an individual basis within each letter environment.

\makelabels % this command prints envelope labels on the final page of the document

\begin{document}
\begin{letter}{Editor \\ University of Minnesota Press \\ Minneapolis, MN}

\opening{Dear Editor,} % eg Hello.

I write to enquire whether you might consider a critical edition of essays by Hugo Gernsback I am currently in the process of editing titled The Perversity of Things: Writings on Media, Technology, and Science Fiction.  This project seeks to patch a major hole in our understanding of early twentieth century media culture and the emergence of science fiction.  [another sentence or two summarizing the book]

My awards/qualifications to write the book?

The primary audience for the Perversity of Things will be…  I have previously published/presented work from this book…  The book will be \# words of text, \# including notes and bibliography.  [something concerning page layout, and see attached mockup]

The manuscript will be ready for submission within six months.  Minnesota’s … series seems a strong fit for my work, given your commitment in recent publications to…

Would you be interested in seeing a fuller description of the project?  I enclose my CV, a \#-page description of the manuscript, a table of contents, and several mock-ups of the page layout I imagine for the book.  I look forward to hearing back from you.

\closing{Sincerely,} % eg Regards,

\cc{} % people this letter is cc-ed to
\encl{} % list of anything enclosed
\ps{} % any post scriptums. ``PS'' labels must be put in manually

\end{letter}
\end{document}
